The importance of a complete description of dependence between variables for ecology and related fields

### 2019-05-06 Paper.Rmd
finished a round of editing the main text, we are now within Ecol Lett length limits

<!--***DAN: stopped here 2019 05 03 321pm-->

@@

@@

\noindent Skewness results were generally consistent with hypotheses. For the datasets that
exhibited strong lower-tail (respectively, upper-tail) 
dependence in earlier analyses (namely, the green spruce 
aphid abundance data for lower-tail, and the leaf-curling plum aphid first flight data
for upper-tail, Table 
\ref{tab_multivar_summary}), the skewness, through time, of the spatial average of the 
empirical data was less than (respectively, greater than) a significant fraction (i.e.,
more than $95\%$) of the
surrogate skewness values 
(Fig. \ref{fig_skewness_spearman}A,B). For the datasets that demonstrated a 
more moderate lower-tail dependence in earlier analyses (the 
\emph{Ceratium furca} abundance and the methane data), skewness of the spatial
average showed a non-significant tendency toward being less than surrogate skewnesses
(Fig. \ref{fig_skewness_spearman}C), or was less than a marginally significant 
fraction of surrogate skewnesses (Fig. \ref{fig_skewness_spearman}D).

As hypothesized, left-tail dependent environmental fluctuations increased metapopulation
extinction risk for the spatially 
extended Lewontin-Cohen model (Fig. \ref{fig_extrisk_lcohen_local}). 
This was true for both $N=5$ and $N=25$ habitat patches, and for both local 
(Fig. \ref{fig_extrisk_lcohen_local}) and global (Fig. \ref{SM-fig_extrisk_lcohen_global})
dispersal. We also examined the influence of non-N copula structure and tail dependence
on Taylor's law in Appendices \ref{SM-Consequence_TL_simulation} and 
\ref{SM-sect:copsurrognd}.

@@

<!--what we did and why is it important to study?-->
We have shown that non-N copula structure and asymmetric tail dependence are 
exceedingly common in datasets that range across multiple sub-disciplines
and levels of organization in ecology, in spite of the fact that no published 
ecological research we are aware of has 
taken this structure into account. Via simple mathematical models, we hypothesized two 
mechanisms that may cause non-N copula structures with asymmetric tail dependence
in ecological data. One
mechanism was via a Moran-like effect by which the dependence structure between ecological
variables (measured through time at different locations) is shaped by the dependence structure of 
environmental variables (measured at the same times and locations). 
A second mechanism gives rise to non-N copula structure and asymmetric tail dependence via
the asymmetric or nonlinear influence of environmental variables on ecological variables, 
and relates to Liebig's law of the minimum. We discuss below how commonly these and other
potential mechanisms may opperate. 
We also demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology. First, the skewness, through time, 
of the spatial average of time series from different locations (e.g., average regional pest abundance computed from local pest abundances) is affected by the tail dependence of the relationships
between the local quantities. Thus right-tail-dependent local quantities 
tend to produce "spiky" spatially averaged time series, with large outbreaks. 
Second, metapopulation extinction risk can be accentuated or reduced by the tail dependence
properties of environmental variables that influence the populations. Finally, 
simulation results in Appendix \ref{SM-Consequence_TL_simulation} 
showed that spatial Taylor's law can be strongly influenced by
tail dependence between local populations. In our view, our results demonstrate that, for many
ecological phenomena, complete understanding is impossible without 
characterizing dependence between variables using copulas.

<!--A para about what is available in the copula and VineCopula packages. Maybe make it a broader para about emthods, and state in there that we developed some new surrogate methods that may be broadly useful for researches trying to incorporate copula approaches.-->
Copula approaches are
well developed, statistically [@joe2014_dependence; @nelsen2006_copula], and 
computer implementations exist (e.g., the `copula` and 
`VineCopula` packages in R). Ecologists need only become aware of these tools
and understand them conceptually to start applying them broadly. 
We created a variety of interrelated randomization procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that build upon existing methods and may be especially useful for ecological applications.

<!--some mechanism paragraphs follow-->

<!--Maybe a para on how likely one is to get copula structure via the Moran mechanism - considers
work on copula structure in environmental variables. Maybe mention possible future work: 
study copula structure of environmental variables (such as temperature and precip) that are known 
to be important in ecology, including how such copula structure may be changing and whether it 
has a geography of the kind studied in Walter et al.-->
One may expect our Moran-like mechanism for non-N copula structure and tail dependence to 
operate commonly, because Moran effects are common [@Liebhold2004], and 
non-N and tail-dependent copula structures have often been found in prior analyses of 
environmental variables. A recent study [@agliardi2019relationship] found non-N 
copulas (mainly Gumbel and Survival Clayton) were better fit in modelling the 
upper tail dependence between extreme large values for global 
temperature and annual mean green house gas concentrations for an extensively
long datasets (1880-2017). If large meteorological events are also spatially widespread, then
one would expect environmental variables associated with these events to take extreme
values simultaneously across large spatial areas, producing tail dependence in measurements
made at different locations. Non-extreme values, not associated with extreme events, may
be more associated with local weather phenomena, and therefore 
less prone to being correlated across large areas.
Using a bivariate mixed model approach with copulas, 
@Serinaldi2008 examined the spatial dependence 
of rainfall observations (7 years with 30 minute time resolution) 
within a rain gauge network in Central Italy. Gumbel or Student 2-copulas 
were possible candidates for modelling bivariate dependence at several time scales, and 
neither of these is an N copula. @liu2016probabilistic used a copula 
based approach in assessing the drought risk with joint probability distribution 
between vegetation and climatic variables across China. 
In a majority of their cases, they found 
non-N copula structure in the dependence between a vegatation index 
and climatic variables (e.g., precipitation and temperature). 
Another long-term study (1950–2014) in the Loess Plateau of China [@She2018]
showed that a Gumbel copula (which has upper tail dependence) could effectively model 
the spatial dependence of drought variables 
(duration and severity) in the study area. Bivariate copula analysis was also used in forecasting the 
co-occurrence of extreme events (flood or drought) 
over the North Sikkim Himalayas, India, using spatial data sets at different 
time scales [@Goswami2018].

<!--Maybe a para on how likely one is to get copula structure via the Liebig mechanism - considers
what people have written on nonlinear effects and empirical evidence for Liebig's law and so on-->
There are also good reasons to expect that our second mechanism operates commonly. 
The second mechanism is the one that involves asymmetric or nonlinear influences of 
environmental variables. Liebig's law of the minimum and the idea of a "limiting 
nutrient" of organism or population growth are fundamental and dominant paradigms in ecology.
Furthermore, innumerable studies document nonlinear influences of environmental
variables on important ecological quantities at all levels of organization, e.g., the 
influence of temperature on: 
individual metabolism and autotroph growth rates 
[@GarciaCarreras2018]; dispersal [@Oconnor2007]; population growth 
[@Savage2004]; and microbial and ecosystem methane flux [@YvonDurocher2014].
Environmental variables influence not only ecological populations, but also physiological,
community, and ecosystem functioning measurements. So our two mechanism may well operate
to produce spatial copula structure in a wide variety of ecological measurements.

<!--Holder stuff-->
Our simulations of character evolution suggest the hypothesis that evolutionary
changes through time in bird and mammal BMR and body size may exhibit right-tail dependence. This is
a hypothesis only, because the right-tail dependence of Fig. \ref{fig_biv_multi_raw_cop}F,G
could have come about in another, unknown way: our simulations show that tail dependence 
in evolutionary changes is sufficient, but may not be necessary, to produce tail
dependence in characters of extant species. For instance, as discussed below and in
Appendix \ref{SM-sect:missingdata} (see also Fig. \ref{SM-fig:missingdata}), 
systematically missing data can produce tail dependence and may
have influenced results for the BMR versus body mass datasets. Even if the hypothesized 
evolutionary mechanism
is correct, our results simply replace one question, i.e., why do we see right-tail
dependence between BMR and body mass for birds and mammals, with another question,
i.e., why might we see right-tail dependence in evolutionary changes of these traits?
See Appendix \ref{SM-evol_mech} for additional thoughts on the potential importance of 
copulas for studies of character evolution when phylogeny branch lengths are not available.

<!--Discuss stuff here on tradeoff and life history that Bever told me about-->
The relationships we studied between BMR and body mass relate
to a trade-off between mass-specific BMR (i.e., BMR per unit body mass) and body
mass itself. Trade-offs between life history characteristics are a focal topic in 
evolutionary ecology. There are reasons to believe copulas interrelate with 
life-history theory in additional ways beyond what we could demonstrate 
with the data we used.
It is well known that energy allocation to a life function, X (e.g., reproduction)
will reduce the energy that can be allocated to other functions, $Y_1$, $Y_2$, $Y_3$ 
(e.g., growth, predation avoidance; this is sometimes called the principle of 
allocation).
But it is equally well known that X can trade-off against any or all of 
the $Y_i$. Therefore, for large values of X which approach absolute 
limitations, there may be a strong 
association between X and $Y_1$. Whereas for small values of X there may be
little association - resources not allocated to X can instead be allocated 
to $Y_1$ or $Y_2$ or to some combination. This constitutes 
asymmetric tail dependence between X and $Y_1$. For instance, @Winemiller1992
described a three-way trade-off in North American fishes between age of reproductive
maturity, juvenile survivorship, and fecundity. The trade-off should, in theory,
produce a tight association between age of maturity and fecundity for 
fishes with low age of maturity, but little association between those
variables for later-maturing
fishes because those fishes may invest much or little in juvenile survival.
Although copulas may be very useful for studying multi-dimensional trade-offs, 
applications will 
require careful attention to the possible consequences of biased sampling:
if the degree of completeness of a dataset is associated with one or
more of the quantities being studied then statistical artefacts
may bias conclusions of copula analyses (Appendix \ref{SM-sect:missingdata}, 
Fig. \ref{SM-fig:missingdata}).

<!-- We have only given a few mechanisms, but there are likely to be others, and discoverig
additional mechanisms of copula structure is certainly worthy of future work. -->
Other mechanisms that we have not considered probably also operate to produce copula structure
in ecological and evolutionary datasets, and should be considered in future work. For instance, 
measurement error may modify copula structure, though one would have to have information
on the precise nature of measurement error in a dataset to explore its influence. 
Not only did the models we used to explore mechanisms not consider measurement error, 
they were intentionally simple in other respects, as well, not including phenomena such as 
delayed density dependence, dispersal, stage structure in population dynamics, and trophic
interactions.
We hope by enumerating a few potential mechanisms of copula structure in data 
we help inspire additional research on the
potentially numerous mechanisms that may operate in diverse datasets.

@@

<!--a para on stability of the total biomass time series in a community, and skewness - sets up Shya's later paper as future work-->
Our results showed that the skewness of the 
spatial average of local time series from different locations is influenced 
by their tail dependence. But the logic of this result should also apply to any collection of time
series, whether associated with different locations in space or not, that are summed or averaged. 
Another potential application is time series of all species from a single community, e.g., all
plants in a quadrat or survey site (REF) or all herbivorous plankton in a lake (REF). A huge 
literature has focused on synchrony versus compensatory dynamics between such time series (REFS), and 
the influence of such interspecific relationships on the variability, through time, of total-community
properties such as total biomass. Typically, variability through time of total-community
properties is measured with the coefficient of variation, but the skewness of such a time series
may also be of interest: it can characterize the extent to
which community properties show "spikiness," i.e., uncharacteristically extreme values. 
Therefore, future work on copula structure of intraspecific relationships in communities and its
influence on community properties seems likely to us to be a potentially very exciting avenue.

<!--A para that sets up Shya's extinction risk paper-->
Although we have demonstrated that tail dependence between environmental variables
can influence extinction risk, substantial work remains to be done regarding the 
circumstances under which this result holds, and its magnitude in different scenarios. 
First, we used 
a non-density dependent model to examine extinction risk. Does tail dependence
in environmental variables still influence extinction risk when density dependence
plays a role? Second, we considered metapopulation extinction risk, but the large field 
of population viability analysis (PVA) via stochastic 
matrix modelling [@Caswell2000; @MorrisDoak2002]
uses a framework in which a single population's vital rates (e.g., life-stage-specific 
fecundity and survival rates) are considered to
vary stochastically through time due to environmental variation. Do relationships
between different vital rates exhibit tail dependence, and does tail dependence 
influence extinction risk of a single population?
PVA is one of the routes by which a species can be added to the influential
International Union for the Conservation of Nature (IUCN) red list of threatened
species. It seems worthwhile for future research to investigate the extent to which 
copula structure in dependences between vital rates may occur and may influence
estimates of risk. Finally, is the copula structure of environmental variables
changing through time, and, if so, how do such changes influence extinction risks?
Climate change is known to amplify the factors that lead to extreme weather events
[@Hansen2012]. Extreme events, by definition, produce values in the right tails of distributions
of environmental variables. Extreme weather events also tend to be spatially extensive, so
increased frequency of extreme weather events may correspond to changes in the tail dependence
environmental variables measured in different places, and we showed such changes
can affect extinction risk. It seems
important for future work to explore whether such effects are of sufficient 
commonness and magnitude to interest conservationists.

>>>>

@@

\noindent Skewness results were generally consistent with hypotheses. For the datasets that
exhibited strong lower-tail (respectively, upper-tail) 
dependence in earlier analyses (namely, the green spruce 
aphid abundance data for lower-tail, and the leaf-curling plum aphid first flight data
for upper-tail, Table 
\ref{tab_multivar_summary}), the skewness of the spatial average 
was less than (respectively, greater than) a significant fraction of the
surrogate skewness values 
(Fig. \ref{fig_skewness_spearman}A,B). For the datasets with moderate lower-tail 
dependence (\emph{Ceratium furca} abundance and methane data), the skewness of the spatial
average showed a non-significant or marginally significant tendency 
toward being less than surrogate skewnesses (Fig. \ref{fig_skewness_spearman}C,D).

@@

As hypothesized, left-tail dependent environmental fluctuations increased metapopulation
extinction risk for the spatially 
extended Lewontin-Cohen model (Fig. \ref{fig_extrisk_lcohen_local}). 
This was true for both $N=5$ and $N=25$ habitat patches, and for both local 
(Fig. \ref{fig_extrisk_lcohen_local}) and global (Fig. \ref{SM-fig_extrisk_lcohen_global})
dispersal. We also showed (Appendix \ref{SM-Consequence_TL_simulation}, \ref{SM-sect:copsurrognd})
that Taylor's law can be strongly influenced by
tail dependence between populations.

@@

<!--what we did and why is it important to study?-->
We showed that non-N copula structure and asymmetric tail dependence are 
exceedingly common in datasets across multiple sub-disciplines
and levels of organization in ecology, although no published 
research we are aware of has 
taken this structure into account. Via simple models, we hypothesized three 
mechanisms that may cause non-N copula structures. One
mechanism was via a Moran-like effect, whereby dependencies between environmental variables 
beget similar dependencies between ecological variables in different locations.
A second mechanism produces tail dependence via nonlinear environmental influences
on ecological variables. We discuss below how commonly these and other
potential mechanisms may opperate. 
We also demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology. First, the skewness
of a spatial-average time series is influenced by tail dependencies between its 
constituent time series. Thus right-tail-dependent local time series 
lead to "spiky" spatially averaged time series, with large outbreaks.
These observations could have implications for numerous exploited and pest species,
for which the spatial average time series have applied importance.
Second, metapopulation extinction risk can be accentuated or reduced by tail dependence
properties of environmental variables. We also showed effects of copula 
structure on Taylor's law, which is has been applied in many fields.
In our view, our results demonstrate that comprehensive understanding 
of many ecological phenomena is impossible without 
a complete characterization of dependence between variables. 
Copula approaches are
well developed, statistically [@joe2014_dependence; @nelsen2006_copula], and 
computer implementations exist (e.g., the `copula` and 
`VineCopula` packages in R); ecologists can start applying these tools
immediately. We created a variety of interrelated randomization procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that build upon existing methods and may be especially useful for ecological applications.

<!--para on how likely one is to get copula structure via the Moran mechanism-->
Our Moran-like mechanism for copula structure and tail dependence seems likely to
operate commonly, because Moran effects are common [@Liebhold2004], and 
non-N and tail-dependent copula structures are often found in 
environmental variables. 
If large meteorological events are 
also spatially widespread, then
one would expect environmental variables associated with these events to take extreme
values simultaneously across large spatial areas, producing tail dependence in measurements
made through time at different locations. Non-extreme values, not associated with extreme, 
spatially extensive weather events, may
be more associated with local phenomena, and therefore 
less prone to being correlated across large areas.
For instance, @Serinaldi2008 examined the spatial dependence 
of rainfall observations in Central Italy. G or Student 2-copulas 
were possible candidates for modelling dependence, and 
neither of these is an N copula.
A long-term study (1950–2014) in the Loess Plateau of China [@She2018]
showed that a G copula (which has upper tail dependence) could effectively model 
the spatial dependence of drought variables in the study area.
Bivariate copula analysis was also used in forecasting the 
co-occurrence of extreme events (flood or drought) 
over the North Sikkim Himalayas, India, using spatial data sets [@Goswami2018].

<!--a para on how likely one is to get copula structure via the Liebig mechanism-->
There are also reasons to expect that our second mechanism operates commonly. 
The second mechanism is the one that involves asymmetric or nonlinear influences of 
environmental variables. Liebig's law and the idea of a "limiting 
nutrient" of growth are dominant paradigms in ecology.
Furthermore, innumerable studies document nonlinear influences of environmental
variables on ecological quantities at all levels of organization.

<!--Holder stuff-->
Our simulations of character evolution suggest the hypothesis that evolutionary
changes through time in bird and mammal BMR and body size may exhibit right-tail dependence,
contrary to standard assumptions of models of character evolution. This is
a hypothesis only, because the right-tail dependence of Fig. \ref{fig_biv_multi_raw_cop}F,G
could have come about in another, unknown way: our simulations show that tail dependence 
in evolutionary changes is sufficient, but may not be necessary, to produce tail
dependence in characters of extant species. For instance, as discussed below and in
Appendix \ref{SM-sect:missingdata} (also Fig. \ref{SM-fig:missingdata}), 
systematically missing data can produce tail dependence and may
have influenced results for the BMR-body mass datasets. Even if the hypothesized 
evolutionary mechanism
is correct, our results simply replace one question, i.e., why do we see right-tail
dependence between BMR and body mass for birds and mammals, with another question,
i.e., why might we see right-tail dependence in evolutionary changes in these traits?
See Appendix \ref{SM-evol_mech} for additional thoughts on the potential importance of 
copulas for studies of character evolution.

<!--Discuss stuff here on tradeoff and life history that Bever told me about-->
The relationships we studied between BMR and body mass relate
to a trade-off between mass-specific BMR (i.e., BMR per unit body mass) and body
mass itself. Trade-offs between life history characteristics are a focal topic in 
ecology. Copulas probably interrelate with 
life-history theory in additional ways beyond what we could demonstrate 
with the data we used.
It is well known that energy allocation to a life function, X (e.g., reproduction)
will reduce the energy that can be allocated to other functions, $Y_1$, $Y_2$, $Y_3$ 
(e.g., growth, predation avoidance; this is called the principle of 
allocation).
But it is equally well known that X can trade-off against any or all of 
the $Y_i$. Therefore, for large values of X which approach absolute 
limitations, there may be a strong 
association between X and $Y_1$. Whereas for small values of X there may be
little association - resources not allocated to X can instead be allocated 
to $Y_1$, $Y_2$, $Y_3$ or to some combination. This constitutes 
asymmetric tail dependence between X and $Y_1$. For instance, @Winemiller1992
described a three-way trade-off in North American fishes between age of reproductive
maturity, juvenile survivorship, and fecundity. The trade-off should, in theory,
produce a tight association between age of maturity and fecundity for 
fishes with low age of maturity, but little association between those
variables for later-maturing
fishes because those fishes may invest much or little in juvenile survival.
Although copulas may be very useful for studying multi-dimensional trade-offs, 
applications will 
require careful attention to the possible consequences of biased sampling:
if the degree of completeness of a dataset is associated with one or
more of the characters being studied then statistical artefacts
may bias conclusions (Appendix \ref{SM-sect:missingdata}, 
Fig. \ref{SM-fig:missingdata}).

<!-- We have only given a few mechanisms, but there are likely to be others, and discoverig
additional mechanisms of copula structure is certainly worthy of future work. -->
Other mechanisms that we have not considered probably also operate, 
and should be considered in future work. For instance, 
measurement error may modify copula structure. 
Our models were intentionally simple in other respects, as well, not including phenomena such as 
delayed density dependence, dispersal, stage structure in population dynamics, and trophic
interactions.
We hope by enumerating a few potential mechanisms of copula structure,
we inspire additional research on the
potentially numerous mechanisms that may operate in diverse datasets.

@@

<!--***DAN: could remove the next para if you need to save words-->
<!--a para on stability of the total biomass time series in a community, and skewness - sets up Shya's later paper as future work-->
Our results showed that the skewness of the 
spatial average of local time series from different locations is influenced 
by their tail dependence. But the logic of this result should also apply to any collection of time
series, whether associated with different locations in space or not, that are summed or averaged. 
Another potential application is time series of all species from a single community, e.g., all
plants in a quadrat surveyed repeatedly over time. A huge 
literature has focused on synchrony versus 
compensatory dynamics between such time series, and 
the influence of such interspecific relationships on the 
variability, through time, of total-community
properties such as total biomass. Typically, variability through time of total-community
properties is measured with the coefficient of variation, but the skewness of such a time series
may also be of interest because it can characterize the extent to
which community properties show "spikiness".  
Future work on copula structure of interspecific relationships in communities and its
implications for community stability seems likely to be an exciting avenue of research.

<!--***DAN: could remove the next para if you need to save words-->
<!--A para that sets up Shya's extinction risk paper-->
Although we demonstrated that tail dependence between environmental variables
can influence extinction risk, substantial work remains regarding the 
circumstances under which this result holds, and its magnitude in different scenarios. 
First, we used 
a non-density-dependent model to examine extinction risk. Do similar results pertain
when density dependence
plays a role? Second, we considered metapopulation extinction risk, but the large field 
of population viability analysis (PVA) via stochastic 
matrix modelling [@Caswell2000; @MorrisDoak2002]
uses a framework in which a single population's vital rates (e.g., life-stage-specific 
fecundity and survival rates) are considered to
vary stochastically through time due to environmental variation. Do relationships
between different vital rates exhibit tail dependence, and does tail dependence 
influence extinction risk of a single population?
PVA is one of the routes by which a species can be added to the influential
IUCN red list of threatened species. Finally, is the copula structure of environmental variables
changing through time, and, if so, how do such changes influence extinction risks?
Climate change is known to amplify the factors that lead to extreme weather events
[@Hansen2012]. Extreme events, by definition, produce values in the right tails of distributions
of environmental variables. Extreme weather events also tend to be spatially extensive, so
increased frequency of extreme weather events may correspond to changes in the tail dependence
of environmental variables measured in different places.
